A Reimplementation of Python's dict using Simple Tabulation Hashing and Linear Probing

### 2013-04-21 docs/paper_lncs.tex
minor fixes

% first the title is needed
\title{A Reimplementation of Python’s dict using Simple Tabulation Hashing and Linear Probing}

% a short form should be given in case it is too long for the running head
\titlerunning{A Reimplementation of Python’s dict}

@@

\begin{abstract}
Here we implement simple tabulation hashing with linear probing in Python
dictionary based on the recent theoretical result of P\v{a}tra\c{s}cu et al. Although
we achieved significant performance wins on synthetic benchmarks, on real
benchmarks our wins were only on the order of 4\% due to interpreter overhead.
Linear probing with sample tabulation hashing does improve Python performance
and should be looked at as a replacement scheme.  As a whole, the
extra memory bandwidth
and latency required to compute the simple tabulation hash negates some of the
performance improvement, but if they were able to be further mitigated, the new scheme
should provide more than a 4\% improvement, making it even more worthwhile.
\keywords{Hashing, Simple Tabulation}
\end{abstract}

@@

Recently, with the advent of packages like numPy, sciPy and pyPy Python has began to also be used (with questionable merit) as a performance computing language due to its ease of use.

However, because Python is an interpreted language it is difficult to operate on a similar performance level to compiled languages due to the sheer overhead of the interpreter.

In order to attempt to improve the performance of the Python language as a whole we plan to attempt to improve the performance of Python dictionaries.

@@

One of the features that we added to the Python source was a conditional flag
that allowed us to view dictionary statistics (collisions, probes, accesses,
etc.) for a given invocation of the Python binary.  With that modification we
were able to see that this code used 92 lookups in string dictionaries and 24
lookups in general purpose dictionaries for a total of 116 dictionary hits. 116
hits in a 3 line piece of code which never uses dictionaries explicitly is quite
significant.  We explain this extensive dictionary usage below.

@@

In this section we give a brief overview of the key implementation details of
dictionaries in CPython 2.7.3 \cite{python27}.  We should note that Python's dictionary supports
several different data types as keys. However, as we noted earlier, the
dictionaries underlying class instances and modules have only strings as keys.
Therefore optimizing a dictionary which have string-only keys is crucial to the
runtime of any Python program. CPython accommodates for this optimization by
changing its dictionary lookup method as necessary. First when a dictionary is
instantiated, CPython uses a string-only lookup function until a search for
non-string is requested, in this case CPython falls back to a more general
lookup method and uses this general function onward. There are two main
optimizations in the string-only lookup function.  First, since the string
comparisons do not raise any exceptions by design, some of the error checking
logic can been removed. Second, the string-only lookup function can skip the
rich set of comparisons ( $<=$, $>=$, $<$, $>$, $==$, $!=$) which arbitrary data
types can provide since string type does not have these special cases.  As it
can be seen, string-only keys can dominate the runtime of a dictionary and
CPython uses some measures to optimize for this common case. Because of this
importance we predominantly focus on the study of new implementations of
hashes for the string type for the rest of the paper.

The current CPython dictionary implementation uses an iterative XORed polynomial
terms for calculating the hash of the strings. In this implementation the hash
value for sequential strings differ only in the least significant bits, for
instance: hash("6.851") = 7824665118634166871 and hash("6.852") =
7824665118634166868. This behavior gives good results for sequential strings
which is a common case. When the table size is $2^i$ taking the least
significant $i$ bits as the index can populate the table without a lot of
collisions. This approach gives better than random results in this case and is
also simple and fast to compute.

On the other hand, this hashing strategy has the tendency to fill contiguous
slots of the hash table. CPython tries to solve this behavior by utilizing a
good collision resolution strategy. The collision resolution strategy currently
used is open addressing with a custom non-linear probing scheme. In this scheme
CPython uses quadratic probing based on this function: $j = (5j + 1) \bmod 2^i$
where $0<j<2^i-1$. This recurrence alone is not enough for a better behavior
since again it behaves the same as linear-probing for some keys since it scans
the table entries in a fixed order. To overcome this issue CPython also uses the
most significant bits of the hash value with the following code snippet:

\begin{verbatim}
slot = hash;
perturb = hash;
while (<slot is full> && <item in slot does not equal the key>) {
       slot = (5*slot) + 1 + perturb;
       perturb >>= 5;
}
\end{verbatim}

Based on this new method the probe sequence will depend on nearly all the bits
of a hash value. We should note that the value of  perturb shift (currently 5)
is crucial for a good probing behavior.  It should be small enough so that
all the significant bits have an effect on the probe sequence but at the same
time it should be large enough so that in really bad cases the high-order hash
bits can affect the earlier iterations.

CPython initializes its table with 8 slots. The resizing strategy is based on a
load factor invariant. Whenever there are $n$ keys in the dictionary, there must
be at least $3n/2$ slots in the table. This load factor is to keep the table
sparse and put a bound on the number of collisions that can happen. CPython
resizes the size of table whenever the load factor invariant gets violated. It
quadruples the size when there are less than 50,000 keys in the table and
doubles the size otherwise.

Overall, the design of CPython dictionary is simple and is optimized for more
common cases such as sequential strings. Since it utilizes quadratic probing
with perturbation, even if a data type does not provide a good hash function,
the table still gets populated with with short chain lengths.  On the other hand
there are some behaviors which can be improved. For instance there is a bad
cache locality because for probing a slot the table might need to jump to the
other slots which are not immediate neighbors and not necessarily in cache.

@@

P\v{a}tra\c{s}cu et al. \cite{mihai} describe simple tabulation hashing for fixed key widths.  For implementing this new hashing scheme, first we needed to figure out a variant of simple tabulation hashing for variable length strings that does not require a table for every possible index (we could potentially cycle through a fixed set of tables).  Performing table lookups is potentially an expensive operation with lots of cache hits mitigating the better cache behavior of linear probing so we will have to be careful about how to implement this. In the following section we explain the implementation details of our approach to these problems.

@@

       \begin{verbatim}
    while (--len >= 0) {
        register long index = (*p++) | ((len&TABLE_MASK)<<8);
        x = x ^ RAND_TABLE_NAME[index];
    }
       \end{verbatim}

@@

To implement linear probing, we modified Python's dictionary code to start at
the table entry defined by the hash value and then search sequentially until it
finds the item it is looking for or an empty cell.
\subsubsection{Optimizations}

@@

In order to analyze the result of this new dictionary we measured three main
metrics. First, we compared the performance of the simple tabulation hashing
with Python’s current hashing scheme on random and sequential strings. Second, we measured the number of probes and collisions and overall the quality of our dictionary on structured and unstructured inputs. Finally, in this section  we show some of the total time spent in dictionary operations by using a set of general Python benchmarks and some real-world programs.
\subsection{Table Number Tuning}

For simple tabulation hashing with variable length strings, we decided to use a
fixed number of tables $N$ and round robin between them for every byte.  We
found the optimal value of $N$ to be 8 experimentally.  We limited our choices
to powers of 2 so that costly modulo operations could be avoided by using bit
masks Fig.~\ref{fig:tables} shows the performance of different numbers of
tabulation tables with both 1 and 2 byte indices.  Eight hash tables indexing
every byte separately resulted in the best performance.    We believe this is due
to the low entropy caused by fewer tables and too much memory bandwidth used by a higher
number of tables.  The optimal result may also be an artifact of the fact that
we were using on the order of magnitude of thousands to millions of dictionary
keys.  Our optimal result resulted using 256 64
bit table entries in each of 8 tables using 16KB of memory.
 \begin{figure}
   \centering
   \includegraphics[width=4.1in]{tables.pdf}
   \caption{Eight tabulation tables with 1 byte indices provides the best performance for a
   dictionary with 6 million keys}
   \label{fig:tables}
 \end{figure}

@@

\subsection{Benchmarks}
Due to the heavy utilization of dictionaries in Python internals, one of the
valuable benchmarks is to see how the new dictionary performs with them them. We used
PyBench benchmark suite \cite{pybench} for this task. The results can be seen in Table
~\ref{pybenchtable}.  We saw significant improvements of around
20\% in performance of some of the internals such as the special and normal
class attribute lookups and string predicates. However, at the same time some
of the other internals such as CPython exception handler were running around
12\% slower compared to the current implementation. On average a 4.5\% improvement in runtime was measured for the whole benchmark suite.

@@

PyBench showed the performance difference in each of the individual CPython
internals however for a more realistic picture of the performance in real world
programs more extensive benchmarking of the new implementation was needed.
Especially since we wanted to know the frequency of the different CPython
internals in real-world applications and therefore the corresponding runtime
changes. The new implementation was tested on several standard real world
programs such as Tornado web server, Django web framework, html5lib and ten other real-world programs.

@@

Different request rates (up to 2000 requests per second) were tested on this
sample server and their corresponding response rates were calculated using
httperf, a HTTP performance measurement tool. Considering the variance of request rates on different iterations of the test, we could not find any significant improvement of the response rate of the new implementation. We found that on the average performance was similar however the new implementation showed a smaller standard deviation in the response rates.

@@

As mentioned before, another set of the benchmarks \cite{bench_suite} that we utilized was based on different standard real-world Python applications. We ran and timed the performance of the new implementation and CPython 2.7.3 side by side using the Python-Mirrors suite. The performance of more than ten applications is indicated in Fig.~\ref{fig:suite}.

@@

\item Simple tabulation yields slightly faster and more consistent performance:
With a careful implementation simple tabulation can yield increased hash
performance on linear probing work loads. Simple tabulation definitely leads to
shorter chain lengths, however, those shorter chain lengths don’t always
justify the performance cost in all cases.

>>>>

% first the title is needed
\title{A Reimplementation of Python's dict using Simple Tabulation Hashing and Linear Probing}

% a short form should be given in case it is too long for the running head
\titlerunning{A Reimplementation of Python's dict}

@@

\begin{abstract}
Here we implement simple tabulation hashing with linear probing in Python
dictionary based on the recent theoretical result of P\v{a}tra\c{s}cu et al. Although
we achieved significant performance wins on synthetic benchmarks, on real
benchmarks our wins were only on the order of 4\% due to interpreter overhead.
Linear probing with simple tabulation hashing does improve Python performance
and should be looked at as a replacement scheme.  As a whole, the
extra memory bandwidth
and latency required to compute the tabulation hash negates some of the
performance improvements, but if they were able to be further mitigated, the new scheme
should provide more than a 4\% improvement, making it even more worthwhile.
\keywords{Hash Function, Simple Tabulation, Linear Probing, Python Dictionary}
\end{abstract}

@@

Recently, with the advent of packages such as NumPy, SciPy, and PyPy Python has began to also be used (with questionable merit) as a performance computing language due to its ease of use. However, because Python is an interpreted language it is difficult to operate on a similar performance level to compiled languages due to the sheer overhead of the interpreter.

In order to enhance the performance of the Python language as a whole we plan to improve the performance of built-in Python dictionary, an associative array implementation which is used to store (\textit{key, value}) pairs.

@@

One of the features that we added to the Python source was a conditional flag
that allowed us to view dictionary statistics (collisions, probes, accesses,
etc.) for a given invocation of the Python binary.  With that modification we
were able to see that this code used 92 lookups in string dictionaries and 24
lookups in general purpose dictionaries for a total of 116 dictionary hits. 116
hits in a 3-line piece of code which never uses dictionaries explicitly is quite
significant.  We explain this extensive dictionary usage below.

@@

In this section we give a brief overview of the key implementation details of
dictionaries in CPython 2.7.3 \cite{python27}, the default and most-widely used implementation
of Python which is written in C.  We should note that Python's dictionary supports
several different data types as keys. However, as we noted earlier, the
dictionaries underlying class instances and modules have only strings as keys.
Therefore optimizing a dictionary which have string-only keys is crucial to the
runtime of any Python program. CPython accommodates for this optimization by
changing its dictionary lookup method as necessary. First when a dictionary is
instantiated, CPython uses a string-only lookup function until a search for
non-string is requested, in this case CPython falls back to a more general
lookup method and uses this general function onward. There are two main
optimizations in the string-only lookup function.  First, since the string
comparisons do not raise any exceptions by design, some of the error checking
logic can been removed. Second, the string-only lookup function can skip the
rich set of comparisons ( $<=$, $>=$, $<$, $>$, $==$, $!=$) which arbitrary data
types can provide since string type does not have these special cases.  As it
can be seen, string-only keys can dominate the runtime of a dictionary and
CPython uses some measures to optimize for this common case. Because of this
importance we predominantly focus on the study of new implementations of
hashes for the string type for the rest of the paper.

The current CPython dictionary implementation uses an iterative XORed polynomial
terms for calculating the hash of the strings. In this implementation the hash
value for sequential strings differ only in the least significant bits, for
instance: hash(``6.851'') = 7824665118634166871 and hash(``6.852'') =
7824665118634166868. This behavior gives good results for sequential strings
which is a common case. When the table size is $2^i$ taking the least
significant $i$ bits as the index can populate the table without a lot of
collisions. This approach gives better than random results in this case and is
also simple and fast to compute.

On the other hand, this hashing strategy has the tendency to fill contiguous
slots of the hash table. CPython tries to remedy this issue by utilizing a
reliable collision resolution strategy. The collision resolution strategy currently
used is open addressing combined with a custom non-linear probing mechanism. In this scheme
CPython uses quadratic probing based on this function: $j = (5j + 1) \bmod 2^i$
where $0<j<2^i-1$. This recurrence alone is not enough to achieve a faster runtime. 
For a set of keys, this strategy behaves the same as linear-probing because it scans
the table entries in a fixed order. To overcome this problem, CPython also uses the
most significant bits of the hash value by using the following code snippet:

\begin{verbatim}
slot = hash;
perturb = hash;
while (<slot is full> AND <item in slot does not equal the key>) {
       slot = (5 * slot) + 1 + perturb;
       perturb = (perturb >> 5); // Shifted right by 5 bits
}
\end{verbatim}

Based on this new method the probe sequence will depend on nearly all the bits
of a hash value. We should note that the value of  perturb shift (currently 5)
is crucial for a good probing behavior.  It should be small enough so that
all the significant bits have an effect on the probe sequence but at the same
time, it should be large enough so that in collision-prone cases the high-order hash
bits can affect the earlier iterations.

CPython initializes its table with 8 slots. The resizing strategy is based on a
load factor invariant. Whenever there are $n$ keys in the dictionary, there must
be at least $3n/2$ slots in the table. This load factor is to keep the table
sparse and put a bound on the number of collisions that can happen. CPython
resizes a table whenever the load factor invariant gets violated. It
quadruples the size when there are less than 50,000 keys in the table and
doubles the size otherwise.

Overall, the design of CPython dictionary is simple and is optimized for more
common cases such as retrieval of sequential strings. Since it utilizes quadratic probing
with perturbation, even if a data type does not provide a good hash function,
the table still gets populated with short chain lengths. However, there is still more room for 
improvements and optimizations. For instance, the current approach incurs suboptimal cache locality because probing a single
slot might require jumping to the other slots which are not its immediate 
neighbors and consequently, not necessarily stored in the cache.

@@

P\v{a}tra\c{s}cu et al. \cite{mihai} describe simple tabulation hashing for fixed key lengths.  For implementing this new hashing scheme, first we needed to figure out a variant of simple tabulation hashing for variable length strings that does not require a table for every possible index (we could potentially cycle through a fixed set of tables).  Performing table lookups is potentially an expensive operation with lots of cache hits mitigating the better cache behavior of linear probing, therefore we have to be careful about how it will be implemented. In the following section we explain the implementation details of our approach to these problems.

@@

\begin{verbatim}
while (--len >= 0) {
  register long index = (*p++) | ((len & TABLE_MASK) << 8);
  x = x ^ RAND_TABLE_NAME[index];
}
\end{verbatim}

@@

To implement linear probing, we modified Python's dictionary code to start at
the table entry defined by the hash value and then search sequentially until it
finds the item it is looking for or an empty cell.

\subsubsection{Optimizations}

@@

In order to analyze the result of this new dictionary we measured three main
metrics. First, we compared the performance of the simple tabulation hashing
with Python's current hashing scheme on random and sequential strings. Second, we measured the number of probes and collisions and overall the quality of our dictionary on structured and unstructured inputs. Finally, in this section  we show some of the total time spent in dictionary operations by using a set of general Python benchmarks and some real-world programs.
\subsection{Table Number Tuning}

For simple tabulation hashing with variable length strings, we decided to use a
fixed number of tables $N$ and round robin between them for every byte.  We
found the optimal value of $N$ to be 8 experimentally.  We limited our choices
to powers of 2 so that costly modulo operations could be avoided by using bit
masks Fig.~\ref{fig:tables} shows the performance of different numbers of
tabulation tables with both 1 and 2 byte indices.  Eight hash tables indexing
every byte separately resulted in the best performance.    We believe this is due
to the low entropy caused by fewer tables and too much memory bandwidth used by a higher
number of tables.  The optimal result may also be an artifact of the fact that
we were using on the order of magnitude of thousands to millions of dictionary
keys.  Our optimal result resulted using 256 64-bit table entries in each of 8 tables using 16KB of memory.
 \begin{figure}
   \centering
   \includegraphics[width=4.1in]{tables.pdf}
   \caption{Eight tabulation tables with 1 byte indices provides the best performance for a
   dictionary with 6 million keys}
   \label{fig:tables}
 \end{figure}

@@

\subsection{Benchmarks}
Due to the heavy utilization of dictionaries in Python internals, one of the
valuable benchmarks is to see how the new dictionary performs with them. We used
PyBench benchmark suite \cite{pybench} for this task. The results can be seen in Table
~\ref{pybenchtable}.  We saw significant improvements of around
20\% in performance of some of the internals such as the special and normal
class attribute lookups and string predicates. However, at the same time some
of the other internals such as CPython exception handler were running around
12\% slower compared to the current implementation. On average a 4.5\% improvement in runtime was measured for the whole benchmark suite.

@@

PyBench showed the performance difference in each of the individual CPython
internals; however, for a more realistic picture of the performance in real world
programs more extensive benchmarking of the new implementation was needed.
Especially since we wanted to know the frequency of the different CPython
internals in real-world applications and therefore the corresponding runtime
changes. The new implementation was tested on several standard real world
programs such as Tornado web server, Django web framework, html5lib and ten other real-world programs.

@@

Different request rates (up to 2000 requests per second) were tested on this
sample server and their corresponding response rates were calculated using
httperf, a HTTP performance measurement tool. Considering the variance of request rates on different iterations of the test, we could not find any significant improvement of the response rate of the new implementation. We found that on the average performance was similar, however, the new implementation showed a smaller standard deviation in the response rates.

@@

As mentioned before, another set of the benchmarks \cite{bench_suite} that we utilized was based on standard real-world Python applications. We ran and timed the performance of the new implementation and CPython 2.7.3 side by side using the Python-Mirrors suite. The performance of more than ten applications is indicated in Fig.~\ref{fig:suite}.

@@

\item Simple tabulation yields slightly faster and more consistent performance:
With a careful implementation simple tabulation can yield increased hash
performance on linear probing work loads. Simple tabulation definitely leads to
shorter chain lengths, however, those shorter chain lengths do not always
justify the performance cost in all cases.
